---
title: "Assignment 10: Data Scraping"
author: "Shiyi Zheng"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
library(tidyverse)
library(rvest)
library(here)
library(lubridate)
here()

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Set the year to 2024
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 

url_durham_2024 <- "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024"
web_durham_2024 <- read_html(url_durham_2024)

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * Contact Person
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "Reginald Hicks", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
Water_system_name <- web_durham_2024 %>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
Water_system_name 

Contact_person <- web_durham_2024 %>%
  html_nodes("div+ table tr:nth-child(4) td:nth-child(2)") %>% 
  html_text()
Contact_person

Ownership <- web_durham_2024 %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
Ownership

MGD_each_month <- web_durham_2024 %>%
  html_nodes("th~ td+ td") %>% 
  html_text() %>%
  as.numeric()
MGD_each_month
length(MGD_each_month)


```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 

df_durham_2024 <- data.frame("Month" = c(1,5,9,2,6,10,3,7,11,4,8,12),
                             "Year" = rep(2024, 12),
                             "Maximum_Day_Use_MGD" = as.numeric(MGD_each_month))

df_durham_2024 <- df_durham_2024 %>%
  mutate(Water_System_Name=!!Water_system_name,
         Contact_Person=!!Contact_person,
         Ownership=!!Ownership,
         Date=my(paste(Month,"-",Year))
         ) %>%
 arrange(Date)
df_durham_2024
#5 

ggplot(df_durham_2024,aes(x=Date,y=Maximum_Day_Use_MGD))+
  geom_line()+
  geom_point()+
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(title = "Water Supply Sources of Max Day Use (MGD) in NC Durham (2024) ",
       x="Month",
       y="Max Day Use (MGD)")
```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs, "PWSID" and "year", that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
#6. 
Scrape_LWSP <- function(the_PWSID, the_year){
the_base_url <- "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid="
the_scrape_url <- paste0(the_base_url, the_PWSID, "&year=", the_year)
print(the_scrape_url)

web_LWSP <- read_html(the_scrape_url)
print(web_LWSP)

Water_system_name_tag <- "div+ table tr:nth-child(1) td:nth-child(2)"
Contact_person_tag <- "div+ table tr:nth-child(4) td:nth-child(2)"
Ownership_tag <- "div+ table tr:nth-child(2) td:nth-child(4)"
MGD_each_month_tag <- "th~ td+ td"

Water_system_name <- web_LWSP%>%
  html_nodes(Water_system_name_tag)%>%
  html_text()%>%
  str_trim()
Contact_person <- web_LWSP %>%
  html_nodes( Contact_person_tag)%>%
  html_text()%>%
  str_trim()
Ownership <- web_LWSP %>%
  html_nodes(Ownership_tag)%>%
  html_text()%>%
  str_trim()
MGD_each_month <- web_LWSP %>%
  html_nodes(MGD_each_month_tag)%>%
  html_text()%>%
  str_trim()%>%
  as.numeric()

df_LWSP <-data.frame("Month" = c(1,5,9,2,6,10,3,7,11,4,8,12),
                     "Year"  = rep(the_year, 12),
                     "PWSID" = rep(the_PWSID, 12),
                     "Maximum_Day_Use_MGD" = as.numeric(MGD_each_month)) %>%
  mutate(Water_System_Name = !!Water_system_name,
      Contact_Person    = !!Contact_person,
      Ownership         = !!Ownership,
      Date = my(paste(Month, "-", Year)))%>%
  arrange(Date)

return(df_LWSP)
}


```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7 
df_Durham_2020 <- Scrape_LWSP('03-32-010',2020)
print(df_Durham_2020)

ggplot(df_Durham_2020,aes(x=Date,y=Maximum_Day_Use_MGD))+
  geom_line()+
  geom_point()+
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(title = "Water Supply Sources of Max Day Use (MGD) in NC Durham (2020) ",
       x="Month",
       y="Max Day Use (MGD)")

str(df_Durham_2020)
glimpse(df_Durham_2020)

```

8. Use the function above to extract data for Cary (PWSID = '03-92-020') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.

```{r fetch.and.plot.Cary.2020.data}
#8 
df_Cary_2020 <- Scrape_LWSP('03-92-020',2020)
print(df_Cary_2020)

ggplot(df_Cary_2020,aes(x=Date,y=Maximum_Day_Use_MGD))+
  geom_line()+
  geom_point()+
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(title = "Water Supply Sources of Max Day Use (MGD) in NC Cary (2020) ",
       x="Month",
       y="Max Day Use (MGD)")

df_Durham_Cary_2020 <- bind_rows(df_Durham_2020,df_Cary_2020)
df_Durham_Cary_2020

ggplot(df_Durham_Cary_2020,aes(x=Date,y=Maximum_Day_Use_MGD,color=Water_System_Name))+
  geom_line()+
  geom_point()+
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(title = "Water Supply Sources of Max Day Use (MGD): Durham vs Cary in NC (2020) ",
       x="Month",
       y="Max Day Use (MGD)")

```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023. Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "06_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs (where one of the two inputs does not change). Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9 

years_range <- c(2018:2023)
PWSID_Cary <- rep('03-92-020', length(years_range))
library(purrr)
df_Cary_2018_2023 <- map2(PWSID_Cary,years_range,Scrape_LWSP)
df_Cary_2018_2023_dataframe <- bind_rows(df_Cary_2018_2023)

ggplot(df_Cary_2018_2023_dataframe,aes(x=Date,y=Maximum_Day_Use_MGD))+
  geom_line()+
  geom_point()+
  geom_smooth(method = "loess", se = FALSE)+
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(title = "Water Supply Sources of Max Day Use (MGD) by month in NC Cary (2018-2023) ",
       x="Year",
       y="Max Day Use (MGD)")

```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?
> Answer:Yes.The smoothed trend line suggests a modest but consistent increase in maximum daily water withdrawals over the 2018–2023 period. While seasonal fluctuations are evident, the long-run pattern indicates gradual growth in water demand.
>
